---
title: "Homework Session I Solutions"
date: today
output:
  html_document: default
  pdf_document: default
---

## Exercise 1
Import the dataset which attached in this folder. 

```{r}
ess <- haven::read_dta(here::here("homework-1/ESS_CZ.dta"))
```


Summarize three variables of your choice which are contained in the dataset. 

Please discuss these three variables which you can see (and write the code to explain how you retrieved these information).

Any abnormalities or something that stands out? Discuss it in your own words!

```{r}
summary(ess)
```


## Exercise 2
Please drop the column containing the variable `prtvtcz` and the column `euftf` of your dataframe. 



Next, find a way to mathematically recode the variable on religious attendance. It is coded backwards so that smaller values correspond to high attendance. 

```{r}
ess <- ess |> 
  select(-c(prtvtcz, euftf)) |>
  mutate(rlgatnd = (8 - rlgatnd)*-1, 
         eduyrs = case_when(
           eduyrs <= 10 ~ eduyrs,
           TRUE ~ NA_integer_
         )) |> 
  slice(51:n())
```


Lastly, find a way to drop the first 50 rows.


### Building a Linear Regression Model
Let’s suppose we are interested in constructing a model that tests the influence of *happiness*, *income* and *education* **on** *left-right self-placement*. 0 on the left-right scale refers to being far-left, whereas 10 indicates that the respondent place themself on the far-right end of the scale. 


```{r}
ess_model <- lm(lrscale ~ happy + hinctnt + eduyrs, data = ess)
```


Build this model in R and display the regression table.

```{r}
library(stargazer)
stargazer(
  ess_model,
  type = "text", 
  star.cutoffs = c(0.05, 0.01, 0.001))
```


Summarize and interpret the coefficients of the model (your text should be written outside of the code chunks!).


```{r}
ess_model_coefficients <- ess_model |> 
  broom::tidy(se = TRUE)

ess_model_coefficients |> 
  filter(term != "(Intercept)") |>
  ggplot(aes(x = term, y = estimate, color = term)) +
  geom_point() +
geom_errorbar(aes(ymin = estimate - 1.96 * std.error, ymax = estimate + 1.96 * std.error), width = 0.2) +
  coord_flip() +
  labs(
    title = "Coefficients of the Linear Regression Model",
    x = "Coefficient",
    y = "Estimate"
  ) +
  theme(
    axis.text.y = element_text(size = 8)
    ) +
  scale_color_brewer(palette = "Set1") +
  geom_hline(yintercept = 0,
             color = "black", 
             linetype = "dashed")
```


Discuss the other values which your regression table shows (R2, Adjusted R2 etc). 


Bonus Points: Plot the coefficients!






